Concurrent Programming in Harmony

### 2021-11-15 book/paper.tex
Updated the appendices to describe "when" statements

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
$e$ & $e$ is an expression\\
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^},
\texttt{and}, \texttt{or}\\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bounded variables, $e$ is a set or dictionary \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{invariant e} & \texttt{e} is an invariant\\
\hline
\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{possibly $b$ [, $b$, ...]} & each $b$ is a boolean \\
\hline
\texttt{select $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
                            \texttt{S} a list of statements\\
\hline
\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
\hline
\texttt{spawn [eternal] m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is the thread-local state \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\end{tabular}

@@

The statement \textbf{atomic $S_1; S_2; ...$} evaluates statements
$S_1, S_2, ...$ atomically in that no other threads can run while
this atomic block is executing.  Typically an atomic block runs
to completion before any other thread can run.  The only exception
to this is if the atomic block executes a \textbf{stop} expression.
In this case, another thread can run.  When the original thread
is resumed (using a \textbf{go} statement), it is once again
atomically executing.

The \textbf{atomic} statement is useful for implementing synchronization
primitives such as test-and-set.  It is also useful for testing.
It is not a replacement for lock/unlock, and should not generally be used
for synchronization otherwise.  Lock/unlock does allow other
threads to run concurrently---just not in the same critical section.

@@

The statement \textbf{await $b$} is equivalent to \textbf{while !$b$: pass}.
It is intended to improve readability of your code.

@@

The statement \textbf{def} \texttt{m} $a: S_1; S_2: ...$
defines a new program counter constant \texttt{m} referring to
a method that takes an argument $a$ and executes the statements
$S_1, S_2, ...$.  The argument $a$ can be a tuple pattern similar
to those used in \texttt{let} and \texttt{for} statements.
Examples include $()$, $(x,)$, $(x, y)$, and $(x, (y, z))$.
The given local variable names variable names are assigned
upon application.
It is allowed, but discouraged, to updates those local variables
in statements $S_1, S_2, ...$.
Each method has a predefined local variable
\textit{result}, initialized to \texttt{None},
that is returned by the method.  Harmony does
not support a \texttt{return} statement that \emph{breaks out}
of the code before executing the last statement.

@@

You can introduce new local variables in a method using the
\texttt{let} expression.
The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates $b$, assigns
the result to local variable $a$, and evaluates statements
$S_1, S_2, ...$.  \textbf{let} supports pattern matching, so you can
write \texttt{let} $x, (y, z) = b: S_1; S_2, ...$.  This will only work
if $b$ is a tuple with two elements, the second of which also being a
tuple with two elements---if not, Harmony will report an error.
The variables may be updates in statements $S_1, S_2, ...$.

@@

\subsection*{\textbf{select ... in ... [ where ... ]}}

The statement \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$
requires that $y$ evaluates to a set value.
The statement does the following three things atomically:
\begin{itemize}
\item it waits until $y$ is non-empty;
\item it selects one element of $y$ non-deterministically (using
a \texttt{choose} expression);
\item it executes statements $S_1, S_2, ...$ with the selected element
assigned to local variable $x$.
\end{itemize}
$x$ may be a pattern, like in \texttt{let}, \texttt{for}, and \texttt{def}
statements.
Harmony reports an error if $y$ evaluates to a value that is not a set.
If \textit{waiting} is an unused local variable,
then \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$ is equivalent to

\begin{code}
\begin{tabbing}
XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
\>\textbf{let} $\mathit{waiting} = \mathbf{True}:$ \\
\>\>\textbf{while} \textit{waiting}: \\
\>\>\>\textbf{atomic}: \\
\>\>\>\>\textbf{if} $y$ != \{\}:\\
\>\>\>\>\>\textbf{let} $x = \mathbf{choose}(y)$:\\
\>\>\>\>\>\>$S_1$ \\
\>\>\>\>\>\>$S_2$ \\
\>\>\>\>\>\>$...$ \\
\>\>\>\>\>$\mathit{waiting} = \mathbf{False}$
\end{tabbing}
\end{code}

The optional \textbf{where} clause can be used to filter the set down to the elements
that satisfy the condition following the \textbf{where} keyword.

@@

@@

Each (non-reserved) identifier in a Harmony program refers to either
a constant, a global shared variable, a local variable, or a module.
Constants are declared using \texttt{const} statements.
Those constants are computed at compile-time.

Local variables all declared.  They can be declared in
\texttt{def} statements (i.e., arguments),
\texttt{let} statements, \texttt{for} loops, and \texttt{select} statements.
Also, each method has an implicitly declared
\textit{result} variable, which is initialized to \texttt{None}.
Each thread has a variable called \textit{this} that contains
the thread-local state.
Local variables are tightly scoped and cannot be shared
between threads.
While in theory one method can be declared within another,
they cannot share local variables either.
All other variables are global and must be initialized
before any threads are spawned.

While arguments to a method and variables in \texttt{for} loops
can be modified, we discourage it for improved code readability.

>>>>

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
$e$ & $e$ is an expression\\
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^},
\texttt{and}, \texttt{or}\\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bounded variables, $e$ is a set or dictionary \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{invariant e} & \texttt{e} is an invariant\\
\hline
\texttt{let $a$ = $e$: S} & $a$ is a bounded variable, $e$ is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{possibly $b$ [, $b$, ...]} & each $b$ is a boolean \\
\hline
\texttt{select $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
                            \texttt{S} a list of statements\\
\hline
\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
\hline
\texttt{spawn [eternal] m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is the thread-local state \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{var $v$ = $e$} & $v$ is a new variable, $e$ is an expression\\
\hline
\texttt{when $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{when exists $a$ in $e$: S} & $a$ is a bounded variable, $e$ is an expression\\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\end{tabular}

\begin{itemize}
\item Bounded variables are read-only.
\item A statement can be preceded by the \texttt{atomically} keyword to make
the statement atomic.
\item Multiple \texttt{for} statements can be combined into a single statement.
\item Multiple \texttt{let} and \texttt{when} statements can be combined into
a single statement.
\end{itemize}

@@

A statement can be preceded by the \texttt{atomically} keyword to
make the statement atomic.
The statement \textbf{atomically: $S_1; S_2; ...$} evaluates statements
$S_1, S_2, ...$ atomically.
This means that the statement runs indivisibly---no other thread can
interleave in the atomic statement.  The only exception
to this is if the atomic block executes a \textbf{stop} expression.
In this case, another thread can run.  When the original thread
is resumed (using a \textbf{go} statement), it is once again
atomically executing.

\textbf{atomically} statements are useful for specification and
implementing synchronization primitives such as test-and-set.
It is also useful for testing.
It is not a replacement for lock/unlock, and should not generally be used
for synchronization otherwise.  Lock/unlock does allow other
threads to run concurrently---just not in the same critical section.

@@

The statement \textbf{await $b$} is equivalent to \textbf{when $b$: pass}.
It is intended to improve readability of your code.

@@

The statement \textbf{def} \texttt{m} $a: S_1; S_2: ...$
defines a new program counter constant \texttt{m} referring to
a method that takes an argument $a$ and executes the statements
$S_1, S_2, ...$.  The argument $a$ can be a tuple pattern similar
to those used in \texttt{let} and \texttt{for} statements.
Examples include $()$, $(x,)$, $(x, y)$, and $(x, (y, z))$.
The given local variable names are assigned upon application and
are read-only.
Each method has a predefined local variable
\textit{result}, initialized to \texttt{None},
that is returned by the method.  Harmony does
not support a \texttt{return} statement that \emph{breaks out}
of the code before executing the last statement.

@@

You can introduce new bound variables in a method using the
\texttt{let} statement.
The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates $b$, assigns
the result to read-only variable $a$, and evaluates statements
$S_1, S_2, ...$.  \textbf{let} supports pattern matching, so you can
write \texttt{let} $x, (y, z) = b: S_1; S_2, ...$.  This will only work
if $b$ is a tuple with two elements, the second of which also being a
tuple with two elements---if not, Harmony will report an error.

@@

@@

\subsection*{\textbf{var}}

You can introduce new local variables in a method using the
\texttt{var} statement.
The statement \textbf{var} $a = b$ evaluates $b$ and assigns
the result to local variable $a$.
\textbf{var} supports pattern matching, so you can
write \texttt{var} $x, (y, z) = b$.  This will only work
if $b$ is a tuple with two elements, the second of which also being a
tuple with two elements---if not, Harmony will report an error.

\subsection*{\textbf{when}}

The statement \textbf{when} $c: S_1; S_2; ...$ executes statements
$S_1, S_2, ...$ after waiting until $c$ evaluates to \textbf{True}.
\texttt{when} statements are most useful when combined with the
\texttt{atomically} keyword.
If \textit{waiting} is an unused local variable,
then \textbf{atomically when} $c: S_1; S_2; ...$ is equivalent to

\begin{code}
\begin{tabbing}
XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
\>\textbf{var} $\mathit{waiting} = \mathbf{True}$ \\
\>\textbf{while} \textit{waiting}: \\
\>\>\textbf{atomically}: \\
\>\>\>\textbf{if} $c$:\\
\>\>\>\>$S_1$ \\
\>\>\>\>$S_2$ \\
\>\>\>\>$...$ \\
\>\>\>\>$\mathit{waiting} = \mathbf{False}$
\end{tabbing}
\end{code}

Multiple \texttt{let} and \texttt{when} statements can be combined.
The expressions before the colon are re-evaluated repeated until all
\texttt{when} conditions are satisfied.

\subsection*{\textbf{when exists ... in ...}}

The statement \textbf{when exists} $x$ \textbf{in} $y: S_1; S_2; ...$
requires that $y$ evaluates to a set value.
The statement does the following three things:
\begin{itemize}
\item it waits until $y$ is non-empty;
\item it selects one element of $y$ non-deterministically (using
a \texttt{choose} expression);
\item it executes statements $S_1, S_2, ...$ with the selected element
bound to read-only variable $x$.
\end{itemize}
$x$ may be a pattern, like in \texttt{let}, \texttt{for}, and \texttt{def}
statements.
Harmony reports an error if $y$ evaluates to a value that is not a set.

\texttt{when} statements are most useful when combined with the
\texttt{atomically} keyword.
If \textit{waiting} is an unused local variable,
then \textbf{atomically when exists} $x$ \textbf{in} $y: S_1; S_2; ...$ is equivalent to

\begin{code}
\begin{tabbing}
XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
\>\textbf{var} $\mathit{waiting} = \mathbf{True}:$ \\
\>\textbf{while} \textit{waiting}: \\
\>\>\textbf{atomic}: \\
\>\>\>\textbf{if} $y$ != \{\}:\\
\>\>\>\>\textbf{let} $x = \mathbf{choose}(y)$:\\
\>\>\>\>\>$S_1$ \\
\>\>\>\>\>$S_2$ \\
\>\>\>\>\>$...$ \\
\>\>\>\>$\mathit{waiting} = \mathbf{False}$
\end{tabbing}
\end{code}

@@

Each (non-reserved) identifier in a Harmony program refers to either
a global constant, a global shared variable, a local bound variable,
a local mutable variable, or a module.
Constants are declared using \texttt{const} statements.
Those constants are computed at compile-time.

Local variables all declared.
Bound variables can be declared in
\texttt{def} statements (i.e., arguments),
\texttt{let} statements, \texttt{for} loops, and \texttt{when exists} statements.
Mutable variables can be declared with the \texttt{var} statement.
Also, each method has an implicitly declared
\textit{result} variable, which is initialized to \texttt{None}.
Each thread has a mutable variable called \textit{this} that contains
the thread-local state.
Local variables are tightly scoped and cannot be shared
between threads.
While in theory one method can be declared within another,
they cannot share local variables either.
All other variables are global and must be initialized
before any threads are spawned.
